---
title: "Assignment 2: Coding Basics"
author: "Michael Gaffney"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1.create sequence of numbers from 1 to 100, skipping by four
numbers <- seq(1,100,4)

#2. set mean and median of the sequence to variables
a <- mean(numbers)
b <- median(numbers)
#3. evaluate variables with a boolean
a > b
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#character vector of student names
students <- c("Matthew", "Mark", "Luke", "John")
#numerical vector of scores
scores <- c(95, 76, 49, 85)
#logical vector of pass fail
passed <- c(TRUE, TRUE, FALSE, TRUE)
#create grades dataframe
grades <- data.frame("name" = students, "score" = scores, "pass/fail" = passed)
print(grades)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: 
They're similiar, but a matrix can only store one type of data (like numeric, for example), where a dataframe can store multiple forms of data.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
passing <- function(x){
  y <- ifelse(x>=50,TRUE,FALSE)
  print(y)
}

passing(grades$score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
I used ifelse becuase the outcome I needed for the scores was binary--if it's above 50, then it's true, and if it's below, it's false. The ifelse function is the simplest and most efficient way to create this output. 
