---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Michael Gaffney"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
library(tidyverse)
library(agricolae)
library(lubridate)
setwd("/Users/michaelgaffney/Documents/Duke University/Nicholas School of the Environment/05 Spring 2022/ENVIRON 872 Environmental Data Analytics/Environmental_Data_Analytics_2022")
lter.data <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
#set the date column using lubridate
lter.data$sampledate <- mdy(lter.data$sampledate)

#2 set the classic themes
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: The null hypothesis is that there is no difference in mean lake temperature with depth for all lakes in July.
Ha: The alternative hypothesis is that there is a difference in mean lake temperature across depth for all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
lter.data.july <- lter.data %>% 
  #keep only July using lubridate function
  filter(month(lter.data$sampledate) == 7) %>% 
  #select desired columns
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  #remove nas
  na.omit()

#create scatterplot of data
ggplot(lter.data.july, aes(x=depth, y=temperature_C)) +
  geom_point() +
  geom_smooth(method=lm) +
  ylim(0,35) +
  ylab("Temperature (C)") + xlab("Depth(m)") +
  labs(title = "Lake Temperature and Depth in July")
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Overall, temperature decreases as depth increases. As we would expect, in other words, the water gets colder as the lake gets deeper. The distribution of points at the greatest depth suggests that temperature's response to depth is nonlinear; i.e., after a certain depth (roughly 10m), the line flattens out and temperature does not decrease proportionally with depth.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lter.july.lm <- lm(data = lter.data.july, temperature_C ~ depth)
summary(lter.july.lm)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: This model shows a highly significant (p < 2.2e-16) negative relationship between temperature and depth. As depth increases, temperature decreases. The model suggests that for every 1m increase in depth, there is a decline in temperature by 1.9 degree C; in other words, for every 1m increase in depth, temperature declines by roughly 2C. At zero depth, the model predicts roughly 22 degrees C as the intercept. On 9726 degrees of freedom, this model predicts .7387 of the variability in the data.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
#run full model
lter.july.AIC <- lm(data = lter.data.july, temperature_C ~ year4 + daynum + depth)
#use stepwise AIC to find the best reduced model
step(lter.july.AIC)

#10 run model suggested by AIC (which happens to be the full model in this case)
lter.july.lm2 <- lm(data = lter.data.july, temperature_C ~ year4 + daynum + depth)
summary(lter.july.lm2)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
While AIC normally takes in a full model and reduces it to a more parsimonious smaller model, in this case the AIC retained the original full model. The variables that explain/predict temperature in the final model are year, date, and depth, with an overall significance at p < 2.2e-16. The model explains .7412 of the variability in the data, which is only a modest improvement on the model that only used depth. This suggests that although year and day number are statistically signficiant, depth might be the single most important explanatory variable.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#ANOVA expressed through the aov function
lter.july.anova.aov <- aov(temperature_C ~ lakename, data = lter.data.july)
summary(lter.july.anova.aov)
#ANOVA expressed as lm function
lter.july.anova.lm <- lm(temperature_C ~ lakename, data = lter.data.july)
summary(lter.july.anova.lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in mean temperature among lakes (p < 2.2e-16). With an adjusted R-Squared of .038 and 9719 degrees of freedom, however, it is clear that variability in lake temperature is explained by many other variables other than the lake itself. It is also clear that not every lake differs in the same way from the other lakes; some lakes are more different than others.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(lter.data.july, aes(x=depth, y=temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method=lm, se = FALSE) +
  ylim(0,35) +
  ylab("Temperature (C)") + xlab("Depth (m)") + 
  labs(title = "Temperature and Depth by Lakes", color = "Lakes")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
#run tukey to explore pairwise difference on the ANOVA model
TukeyHSD(lter.july.anova.aov)
lake.groups <- HSD.test(lter.july.anova.aov, "lakename", group = TRUE)
lake.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake and Paul lake do not have statisitcally different mean temperatures, nor do Ward Lake and Peter Lake. It does not appear from the grouping function that any one lake has a statistically different mean temperautre from every other lake.

 

17. If we were just looking at Peter Lake and Paul Lake, what's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We might also use a two sample T-test, which essenetially performs the same kind of test as a linear model but which uses different math to reach the result. In order to conduct this test, we would need to check whether the variance of the two groups was equivalent.


